Smooth and Persistent Forecasts of German GDP

### 2025-06-21 HWI SSA_27_05_2025_with new poster/22.06.2025/paper.tex
up

The models encapsulate key characteristics of the data after adjusting for publication lags. Notably, GDP exhibits the strongest response to shocks in the two survey indicators—ifo and ESI\footnote{There is also weaker evidence of a response to the term spread at longer lags.}. The impact of the pandemic modifies the lag structure by emphasizing shorter dependencies during that specific episode. Removing the COVID-19 breakout from the analysis results in smoother, somewhat longer-tailed dependencies, consistent with the observed empirical autocorrelation functions. In particular, the VAR(3) model captures slightly extended lead times and more nuanced lag patterns. The delayed response of GDP to movements in these indicators suggests that multivariate models could yield superior forecasting performance compared to univariate approaches, a hypothesis we explore in the next section\footnote{For simplicity, we focus on the VAR(1) model here, while the M-SSA package provides more advanced options, including the discussed VAR(3).}.

@@

For illustration, consider the problem of nowcasting HP-filtered GDP, denoted by HP-GDP, where the target $z_t$ is generated using the two-sided HP(160) filter examined previously. We compare the classic multivariate mean squared error (MSE) predictor (denoted M-MSE) with the M-SSA predictor where we set $HT_1=1.5\times HT_{MSE}$. This constraint ensures that the M-SSA predictor produces roughly 33$\%$ fewer sign changes than M-MSE in large samples. Comparisons with M-MSE allow us to see the cost of this constraint in terms of prediction performance (target correlation). 
Both predictors use the same underlying VAR(1) model introduced in Section \ref{sec:direct_forecast} and estimated on data up to Q4-2007. This ensures a lengthy out-of-sample period that includes significant events such as the financial crisis, the sovereign debt crisis, and the COVID-19 pandemic\footnote{The economic sentiment indicator (ESI) is excluded from the parsimonious VAR(1)-model when using the shorter sub-sample, ending in Q4-2007. Further details and background information are available in \cite{Wildi2025}. All calculations are performed using the M-SSA package, as documented in \cite{Wildi2025}. \url{https://github.com/wiaidp/R-package-SSA-Predictor}}

@@

The second source of revisions in the M-SSA component predictor arises from updates to the VAR model within the M-SSA filter. We compare the original estimates, based on data up to Q4-2007, with the final estimates obtained using $h=4$ and $shift=4$ (similar findings apply to other values of forecast horizon and forward-shift).  For consistency, we excluded the entire pandemic episode from the data. Figure \ref{up_dated_ma_inv_multi_ip} compares the MA-inversions (impulse responses) of the GDP equation derived from the `old' and the updated VAR models: unlike the former, the latter incorporates ESI as an additional explanatory variable for GDP. Figure \ref{bk_2008_all} illustrates the resulting effect on the M-SSA filter targeting HP-GDP at forecast horizon $h=4$. The updated filter (right panel) assigns more weight to the additional `leading' indicators relative to GDP. Finally, Figure \ref{revisions3} compares the M-SSA GDP predictors. To isolate the effect of the regression revision, the figure displays standardized series. The update to the VAR model influences the left-shift of the predictor (blue line), which appears slightly `faster' than the original predictor (red line). This suggests that the out-of-sample performance of the M-SSA component predictor may be better than reported in Section \ref{oosp}, which is based on the fixed and increasingly outdated version of M-SSA.

@@

To address these problems, we propose a multivariate extension of SSA framework based on \cite{Wildi2024}. Unlike the univariate HP filter, the M-SSA can leverage cross-sectional information from indicators leading GDP in real time while controlling for the rate of zero-crossings of the predictor. Consequently, the multivariate filter outputs become increasingly left-shifted as the forecast horizon lengthens, allowing dips and peaks of the target series to be tracked more systematically, while reducing the occurrence of noisy zero-crossings. However, since M-SSA does not explicitly target GDP, an additional step is necessary to derive the final GDP predictor.\\% We propose two approaches: a simple equally-weighted aggregate, assuming all M-SSA components are equally important for predicting GDP, and an optimally weighted predictor based on a regression of the M-SSA components on future GDP.\\

For illustration, we consider a simple approach based on regressing the single M-SSA GDP output on future GDP, thereby ignoring the other filter outputs. This predictor is intuitively appealing because the future HP-GDP, i.e., the target of M-SSA—is the low-frequency component of the future GDP target. Consequently, a strong link between M-SSA and future HP-GDP also indicates a connection with GDP, even if the statistical significance is obscured by the noise inherent in GDP. Out-of-sample performance suggests that this predictor is statistically significant at horizons up to one year. Moreover, designs optimized for larger forecast horizons outperform the mean benchmark at all considered forecast horizons, and the predictor exceeds the performance of direct forecasts---based on unfiltered or HP filtered indicators---at horizons longer than two quarters. An analysis of revision errors arising from quarterly updates of the proposed M-SSA GDP component predictor indicates that the regression parameters, the VAR model, and the resulting M-SSA filters stabilize after an initial burn-in period. This stability enhances the explainability and interpretability of the predictor.\\

>>>>

%The models encapsulate key characteristics of the data after adjusting for publication lags. Notably, 
GDP exhibits the strongest response to shocks in the two survey indicators—ifo and ESI\footnote{There is also weaker evidence of a response to the term spread at longer lags.}. The impact of the pandemic modifies the lag structure by emphasizing shorter dependencies triggered by the alternating outliers during that specific episode. Removing the COVID-19 breakout from the analysis results in smoother, somewhat longer-tailed dependencies, consistent with the observed empirical autocorrelation functions. In particular, the VAR(3) model captures slightly extended lead times and more nuanced lag patterns\footnote{The VAR(3) model yields marginal improvements in multivariate filtering performance. However, due to its relative simplicity, this study primarily employs the VAR(1) model. The M-SSA package (accessible at \url{https://github.com/wiaidp/R-package-SSA-Predictor}) offers more advanced capabilities, including the implementation of the VAR(3) model and various additional (Bayesian) VAR specifications.}. The delayed response of GDP to movements in these indicators suggests that multivariate models could yield superior forecasting performance compared to univariate approaches, a hypothesis we explore in the next section

@@

For illustration, consider the problem of nowcasting HP-filtered GDP, denoted by HP-GDP, where the target $z_t$ is generated using the two-sided HP(160) filter examined previously. We compare the classic multivariate mean squared error (MSE) predictor (denoted M-MSE) with the M-SSA predictor where we set $HT_1=1.5\times HT_{MSE}$. This constraint ensures that the M-SSA predictor produces roughly 33$\%$ fewer sign changes than M-MSE in large samples. Comparisons with M-MSE allow us to see the cost of this constraint in terms of prediction performance (target correlation). 
Both predictors use the same underlying VAR(1) model introduced in Section \ref{sec:direct_forecast} and estimated on data up to Q4-2007. This ensures a lengthy out-of-sample period that includes significant events such as the financial crisis, the sovereign debt crisis, and the COVID-19 pandemic\footnote{In the parsimonious VAR(1) model estimated over the shorter sub-sample ending in Q4-2007, the economic sentiment indicator (ESI) does not receive any weight. Additional technical details and background information are provided in \cite{Wildi2025}. All computations are conducted using the M-SSA package, as documented in \cite{Wildi2025}. The package can be accessed at \url{https://github.com/wiaidp/R-package-SSA-Predictor}.}

@@

The second source of revisions in the M-SSA component predictor arises from updates to the VAR(1) model within the M-SSA filter\footnote{Similar findings apply to more sophisticated alternatives such as the VAR(3) discussed in Section \ref{sec:data}.}. We compare the original estimates, based on data up to Q4-2007, with the final estimates obtained using $h=4$ and $shift=4$ (similar findings apply to other values of forecast horizon and forward-shift).  For consistency, we excluded the entire pandemic episode from the data. Figure \ref{up_dated_ma_inv_multi_ip} compares the MA-inversions (impulse responses) of the GDP equation derived from the `old' and the updated VAR models: unlike the former, the latter incorporates ESI as an additional explanatory variable for GDP. Figure \ref{bk_2008_all} illustrates the resulting effect on the M-SSA filter targeting HP-GDP at forecast horizon $h=4$. The updated filter (right panel) assigns more weight to the additional `leading' indicators relative to GDP. Finally, Figure \ref{revisions3} compares the M-SSA GDP predictors. To isolate the effect of the regression revision, the figure displays standardized series. The update to the VAR model results in a leftward shift of the predictor (represented by the blue line), which appears marginally more rapid compared to the original predictor (depicted by the red line). It is important to note that the smoothness (zero-crossing rate) remains unchanged, though. 
This suggests that the out-of-sample performance of the M-SSA component predictor may be better than reported in Section \ref{oosp}, which is based on the fixed and increasingly outdated version of M-SSA.

@@

To address these problems, we propose a multivariate extension of the SSA framework based on \cite{Wildi2024}. Unlike the univariate HP filter, the M-SSA can leverage cross-sectional information from indicators leading GDP in real time while controlling for the rate of zero-crossings of the predictor. Consequently, the multivariate filter outputs become increasingly left-shifted as the forecast horizon lengthens, allowing dips and peaks of the target series to be tracked more systematically, while reducing the occurrence of noisy zero-crossings. However, since M-SSA does not explicitly target GDP, an additional step is necessary to derive the final GDP predictor.\\% We propose two approaches: a simple equally-weighted aggregate, assuming all M-SSA components are equally important for predicting GDP, and an optimally weighted predictor based on a regression of the M-SSA components on future GDP.\\

For illustration, we consider a simple approach based on regressing the single M-SSA GDP output on future GDP, thereby ignoring the other filter outputs. This predictor is intuitively appealing because the future HP-GDP, i.e., the target of M-SSA—is the low-frequency component of the future GDP target. Consequently, a strong link between M-SSA and future HP-GDP also indicates a connection with GDP, even if the statistical significance is obscured by the noise inherent in GDP. Out-of-sample performance suggests that this predictor is statistically significant at horizons exceeding one year. Moreover, designs optimized for larger forecast horizons outperform the mean benchmark at all considered forecast horizons, and the predictor exceeds the performance of direct forecasts---based on unfiltered or HP filtered indicators---at horizons longer than two quarters. An analysis of hit and false alarm rates within the ROC curve confirms that filtering improves predictive performance for forecast horizons exceeding two quarters. Multivariate approaches can further capitalize on leading indicators within the cross-sectional data. Additionally, the AUC is optimized through more precise control of the zero-crossing rate within the M-SSA framework. Finally, an examination of revision errors associated with quarterly updates of the proposed M-SSA-based GDP component predictor indicates that the regression parameters, VAR model, and resulting M-SSA filters stabilize after an initial burn-in period. This stabilization enhances the model’s explainability and interpretability.\\
